---
title: "R Review"
output: html_notebook
---

For this review, we'll be working with dispatch data from the Lincoln Police Department. You can find the data online here: [http://opendata.lincoln.ne.gov/datasets/lpd-dispatch-records](http://opendata.lincoln.ne.gov/datasets/lpd-dispatch-records).

Use the Download menu to select a csv download.

Then move the data file you downloaded into your project folder.

### Load packages

Load the tidyverse, janitor and lubridate packages.

```{r}
library(tidyverse)
library(janitor)
library(lubridate)
```

### Load data

Load your data into an object called 'dispatches'.

```{r}
dispatches <- read_csv("LPD_Dispatch_Records.csv")
```

#### Fix column names

Use janitor to make all column names comply with R preferred style - all lowercase and underscores between words.

```{r}
dispatches <- dispatches %>% clean_names()
```


### Analysis questions

#### Datatypes

Look at the documentation for the data. Do all the columns appear to be formatted correctly?

Use lubridate to change the RPT_Date column to ymd format.

```{r}
dispatchesdate <- dispatches %>% mutate(
  CleanDate = ymd(rpt_date)
)

dispatchesdate
```

Find the 13 rows that failed to parse. Why?

  I used Summary(dispatch) to find 13 rows had no loc_code and rpt_date had a min of 0 (which shouldn't be possible for a date)

How many cases are in the data? How many unique cases?

There are 2 unique cases with different addresses repeated 4 and 9 times, respectively.



```{r}
filtered <- dispatches %>% filter(rpt_date == "0")

filtered
```

#### Arranging

What are the oldest and most recent cases in the data?

The oldest is from 20170101, AKA January 01, 2017
The newest is from 20200610, AKA June 10, 2020

```{r}
dispatches %>% 
  arrange(rpt_date)

```

#### Filtering

Create a dataframe called 'missing' with just missing persons cases in it.

```{r}
missing <- dispatches %>% filter(cfs_legend == "MISSING PERSON")

missing
```

Use the str_detect function to find all the cases that mention O Street in the address.

```{r}
ostdispatches <- dispatches %>%
  filter(str_detect(block_address, "O ST"))

ostdispatches
```

#### Counting

Use the count() function to find the number of dispatches to each neighborhood. 

```{r}
neighborcount <- dispatches %>% 
  group_by(neighbor) %>%
  summarise(
    count = n()
  ) %>% arrange(desc(count))

neighborcount
```

Which neighborhood appears the most in the data?

Downtown

Do you see any limitations of this data when we are counting by neighborhood?

Yes, most entries in the neighbor column are N.A. That means most often, police aren't trackig the neighborhoods and thus the remaining data is unreliable. 

#### Mutating

Create a new column called 'year' that includes just the year from date_fixed.

```{r}
dispatchesyears <- dispatchesdate %>% mutate(
    rpt_year = year(CleanDate)
  )

dispatchesyears
```

Use the case_when function to create a new categorical variable dividing the rpt_time column into the 24 hours of the day. Be careful to make sure that each time fits into only one value of the new variable.

```{r}
dispatches <- dispatches %>% mutate(
  hour_of_day = case_when(
    rpt_time < 100 ~ 0, 
    rpt_time >= 100 & rpt_time < 200 ~ 1,
    rpt_time >= 200 & rpt_time < 300 ~ 2,
    rpt_time >= 300 & rpt_time < 400 ~ 3, 
    rpt_time >= 400 & rpt_time < 500 ~ 4,
    rpt_time >= 500 & rpt_time < 600 ~ 5,
    rpt_time >= 600 & rpt_time < 700 ~ 6, 
    rpt_time >= 700 & rpt_time < 800 ~ 7,
    rpt_time >= 800 & rpt_time < 900 ~ 8,
    rpt_time >= 900 & rpt_time < 1000 ~ 9,
    rpt_time >= 1000 & rpt_time < 1100 ~ 10, 
    rpt_time >= 1100 & rpt_time < 1200 ~ 11, 
    rpt_time >= 1200 & rpt_time < 1300 ~ 12, 
    rpt_time >= 1300 & rpt_time < 1400 ~ 13,
    rpt_time >= 1400 & rpt_time < 1500 ~ 14, 
    rpt_time >= 1500 & rpt_time < 1600 ~ 15,
    rpt_time >= 1600 & rpt_time < 1700 ~ 16,
    rpt_time >= 1700 & rpt_time < 1800 ~ 17,
    rpt_time >= 1800 & rpt_time < 1900 ~ 18,
    rpt_time >= 1900 & rpt_time < 2000 ~ 19, 
    rpt_time >= 2000 & rpt_time < 2100 ~ 20,
    rpt_time >= 2100 & rpt_time < 2200 ~ 21, 
    rpt_time >= 2200 & rpt_time < 2300 ~ 22,
    rpt_time >= 2300 & rpt_time < 2400 ~ 23
  )
)

```


What hour of the day sees the most police action?

```{r}
caseshours <- dispatches %>%
  group_by(hour_of_day) %>%
  summarise(
    count = n()
  ) %>% arrange(count)

caseshours
```




#### Grouping and summarizing 

How many drug-related dispatches occurred in each year of the data?

```{r}
dispatchesyears %>% group_by(rpt_year) %>%
 filter(str_detect(cfs_legend, "DRUGS"))
```

Create a new column called month. Then using group_by and summarize, find the maximum, minimum and average number of dispatches per month.

```{r}
dispatchesmonth <- dispatchesdate %>% mutate(
    rpt_month = month(CleanDate)
  )

dispatchesmonth %>% arrange(rpt_month)
```
```{r}
monthcount <- dispatchesmonth %>%
  group_by(rpt_month) %>%
  summarise(
    count = n()
  ) %>% arrange(count)

monthcount
```

```{r}
summary(dispatchesmonth)
```



#### Percent change

What was the percent change in total number of dispatches from 2018 to 2019?

82.7%

```{r}
yeartotal <- dispatchesyears %>%
  group_by(rpt_year) %>%
  summarise(
    count = n()
  ) %>% arrange(count)

yeartotal
```

```{r}
percentchange1718<- ((118088-117120)/117120) * 100

percentchange1718
```


#### Line charts

Using ggplot, create a line chart of the number of cases per month and year. Choose an appropriate color for the line, add a title and labels and a choose a theme.

```{r}
library(zoo)

yearmonthtotal <- dispatchesdate %>% 
  mutate(yearmonth = as.yearmon(CleanDate)) %>%
  group_by(yearmonth) %>%
  summarise(count = n()) 

yearmonthtotal
```


```{r}
yearmonthtotal %>%
  ggplot(aes(x = yearmonth, y = count)) +
  geom_line(size = 1) +
  theme_minimal() +
  scale_y_continuous("",labels = scales::comma)
  
```


What do you observe about the yearly pattern of police dispatches? Why do you suppose that is?

#### Column charts

Using ggplot, create a column chart that shows the five most common categories in the cfs_legend column. Apply appropriate design decisions to your chart.

```{r}
casecount <- dispatches %>%
  group_by(cfs_legend) %>%
  summarise(
    count = n()
  ) %>% arrange(count) %>% 
top_n(5, count) -> topcases

casecount
```

```{r}
ggplot() + geom_bar(data=topcases, aes(x=reorder(cfs_legend,count), weight=count)) + coord_flip() +theme_minimal()
```


